---
title: "PhyloAcc accelerated mammal elements"
author: "[gwct](https://gwct.github.io/)"
date: "`r format(Sys.time(), '%m/%d/%Y %H:%M:%S %Z')`"
output:
  #html_document:
  rmdformats::robobook:
    highlight: kate
    includes:
      in_header: '../html-chunks/rmd_header.html'
    df_print: paged
    code_folding: hide
    number_sections: true
    toc_depth: 3
---


```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
library(dplyr)
library(cowplot)
library(phangorn)
library(ape)
library(kableExtra)
library(ggbeeswarm)
library(ggsignif)
library(here)
source("../lib/design.r")
```

[< Back ](index.html)

Comparisons of gene trees for elements in the mammal dataset inferred to be accelerated using the species tree model

```{r read, out.width="100%", fig.align = "center", warning=FALSE}
species_tree_file = here("docs", "data", "mammal.tre")
species_tree = read.tree(file=species_tree_file)

gene_tree_file = here("docs", "data", "mammal-acc-gene-trees-labeled.txt")
gene_trees = read.csv(gene_tree_file, sep="\t", header=F)
names(gene_trees) = c("locus.id", "iqtree.tree.str")

gene_tree_results_file = here("docs", "data", "mammal_gt_model_elem_lik.txt")
elem_lik = read.csv(gene_tree_results_file, sep="\t", header=T)
names(elem_lik)[1] = "run.id"

id_key_file = here("docs", "data", "mammal-id-key.txt")
id_key = read.csv(id_key_file, sep="\t", header=F)
names(id_key) = c("run.id", "locus.id", "max.marginal.lik", "best.model", "phyloacc.tree.str")

data = merge(gene_trees, id_key, by="locus.id")
data = merge(data, elem_lik, by="run.id")

```

# Summary

Elements are said to be accelerated if logBF1 > 5 and logBF2 > 5.

```{r acc-summary, out.width="50%", fig.align = "center", warning=FALSE}

data$acc = "Not accelerated"
data$acc[data$logBF1 > 5 & data$logBF2 > 5] = "Accelerated"

total_elem = nrow(data)
acc_elem = nrow(subset(data, acc=="Accelerated"))
prop_acc_elem = acc_elem / total_elem

table_data = data.frame("Category"=c("Total elements", "Accelerated elements", "Proportion accelerated"), "Value"=c(total_elem, acc_elem, signif(prop_acc_elem, 4)))
#table_data$Value <- sub("0+$", "", as.character(table_data$Value))

table_data %>% kable() %>% kable_styling(bootstrap_options=c("striped", "condended", "responsive"), full_width=F)

```

# Bayes factor and Robinson-Foulds distance to species tree

```{r calc-rf, out.width="80%", fig.align = "center", warning=FALSE, message=FALSE, fig.height=2.5}
data$pa.rf = NA
data$pa.nrf = NA
data$pa.wrf = NA
data$pa.spr = NA
data$pa.kf = NA
data$pa.path = NA
data$pa.wpath = NA
data$pa.num.tips = NA

data$iq.rf = NA
data$iq.nrf = NA
data$iq.wrf = NA
data$iq.spr = NA
data$iq.kf = NA
data$iq.path = NA
data$iq.wpath = NA
data$iq.num.tips = NA

for(i in 1:nrow(data)){
  if(is.na(data[i,]$phyloacc.tree.str)){
    next
  }
  #print(data[i,]$run.id)
  pa_gt = read.tree(text=data[i,]$phyloacc.tree.str)
  # Read the gene tree as a phylo object
  
  cur_tips = pa_gt$tip.label
  pruned_tree = drop.tip(species_tree, species_tree$tip.label[-match(cur_tips, species_tree$tip.label)])
  # Get a list of the tips and prune the species tree to contain only those tips
  
  cur_rf = RF.dist(pruned_tree, pa_gt, normalize=F)
  # Calculate RF between the pruned species tree and the gene tree
  
  cur_nrf = RF.dist(pruned_tree, pa_gt, normalize=T)
  # Calculate normalized RF between the pruned species tree and the gene tree
  
  cur_wrf = wRF.dist(pruned_tree, pa_gt)
  # Calculate weighted RF between the pruned species tree and the gene tree
  
  cur_spr = SPR.dist(pruned_tree, pa_gt)
  
  cur_kf = KF.dist(pruned_tree, pa_gt)
  
  cur_path = path.dist(pruned_tree, pa_gt)
  
  cur_wpath = path.dist(pruned_tree, pa_gt, use.weight=T)
  
  data[i,]$pa.rf = cur_rf
  data[i,]$pa.nrf = cur_nrf
  data[i,]$pa.wrf = cur_wrf
  data[i,]$pa.spr = cur_spr
  data[i,]$pa.kf = cur_kf
  data[i,]$pa.path = cur_path
  data[i,]$pa.wpath = cur_wpath
  data[i,]$pa.num.tips = length(cur_tips)
  # Add the gene tree info to the df, with a special case for when the RF is 0
  
  
  iq_gt = read.tree(text=data[i,]$iqtree.tree.str)
  # Read the gene tree as a phylo object
  
  cur_tips = iq_gt$tip.label
  pruned_tree = drop.tip(species_tree, species_tree$tip.label[-match(cur_tips, species_tree$tip.label)])
  # Get a list of the tips and prune the species tree to contain only those tips
  
  cur_rf = RF.dist(pruned_tree, iq_gt, normalize=F)
  # Calculate RF between the pruned species tree and the gene tree
  
  cur_nrf = RF.dist(pruned_tree, iq_gt, normalize=T)
  # Calculate normalized RF between the pruned species tree and the gene tree
  
  cur_wrf = wRF.dist(pruned_tree, iq_gt)
  # Calculate weighted RF between the pruned species tree and the gene tree
  
  cur_spr = SPR.dist(pruned_tree, iq_gt)
  
  cur_kf = KF.dist(pruned_tree, iq_gt)
  
  cur_path = path.dist(pruned_tree, iq_gt)
  
  cur_wpath = path.dist(pruned_tree, iq_gt, use.weight=T)
  
  data[i,]$iq.rf = cur_rf
  data[i,]$iq.nrf = cur_nrf
  data[i,]$iq.wrf = cur_wrf
  data[i,]$iq.spr = cur_spr
  data[i,]$iq.kf = cur_kf
  data[i,]$iq.path = cur_path
  data[i,]$iq.wpath = cur_wpath
  data[i,]$iq.num.tips = length(cur_tips)
  # Add the gene tree info to the df, with a special case for when the RF is 0
  
  
  
}
# Read all the gene trees

```

## Robinson-Foulds distribution

For each element, a gene tree was inferred with IQ-tree and the Robinson-Foulds distance of that gene tree to the species tree was computed.

### Regular RF

Score is incremented by 1 for each differing branch between the trees

```{r rf-dists, out.width="65%", fig.align = "center", warning=FALSE, fig.height=4}

x_comps = list(c("Accelerated", "Not accelerated"))

p1 = ggplot(data, aes(x=acc, y=pa.rf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("PhyloAcc gene trees") +
  xlab("") +
  ylab("RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p1)

p2 = ggplot(data, aes(x=acc, y=iq.rf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("IQTree gene trees") +
  xlab("") +
  ylab("RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p2)

p = plot_grid(p1, p2, ncol=2)
print(p)

```

### Normalized RF

Score is incremented by 1 for each differing branch between the trees and the final score is divided by the total possible number of differing branches

```{r nrf-dists, out.width="65%", fig.align = "center", warning=FALSE, fig.height=4}

x_comps = list(c("Accelerated", "Not accelerated"))

p1 = ggplot(data, aes(x=acc, y=pa.nrf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("PhyloAcc gene trees") +
  xlab("") +
  ylab("Normalized RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p1)

p2 = ggplot(data, aes(x=acc, y=iq.nrf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("IQTree gene trees") +
  xlab("") +
  ylab("Normalized RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p2)

p = plot_grid(p1, p2, ncol=2)
print(p)

```

### Weighted RF

Score is incremented by the branch length of the differing branch for each differing branch between the trees

```{r wrf-dists, out.width="65%", fig.align = "center", warning=FALSE, fig.height=4}

x_comps = list(c("Accelerated", "Not accelerated"))

p1 = ggplot(subset(data, pa.wrf < 4), aes(x=acc, y=pa.wrf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("PhyloAcc gene trees") +
  xlab("") +
  ylab("Weighted RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p1)

p2 = ggplot(subset(data, iq.wrf < 8), aes(x=acc, y=iq.wrf, group=acc, color=acc)) +
  #geom_boxplot() +
  geom_quasirandom(size=2, width=0.25, alpha=0.25) +
  geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  geom_signif(comparisons=x_comps, map_signif_level=TRUE, textsize=4, size=1, step_increase=0.12, margin_top=0.1, color="#333333") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(limits=c(0,200)) +
  ggtitle("IQTree gene trees") +
  xlab("") +
  ylab("Weighted RF to species tree") +
  bartheme() +
  theme(legend.position="none",
        plot.title=element_text(hjust=0, vjust=6, size=12))
#print(p2)

p = plot_grid(p1, p2, ncol=2)
print(p)

```

## BF1

```{r bf1-rf, out.width="60%", fig.align = "center", warning=FALSE, fig.height=4}

p1 = ggplot(subset(data, logBF1 < 200), aes(x=logBF1, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  xlab("log BF1") +
  ylab("Weighted RF to species tree\n(IQTree gene trees)") +
  bartheme() +
  theme(legend.position="bottom")
#print(p1)

p2 = ggplot(subset(data, logBF1 < 20 & logBF1 > -5), aes(x=logBF1, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  xlab("log BF1") +
  ylab("") +
  bartheme() +
  theme(legend.position="bottom")
#print(p2)

p3 = plot_grid(p1 + theme(legend.position="none"), p2 + theme(legend.position="none"), ncol=2)
pleg = get_legend(p1)
p = plot_grid(p3, pleg, ncol=1, rel_heights=c(1, .1))
print(p)

```

## BF2

```{r bf2-rf, out.width="60%", fig.align = "center", warning=FALSE, fig.height=4}

p1 = ggplot(subset(data, logBF2 > -2), aes(x=logBF2, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  xlab("log BF2") +
  ylab("Weighted RF to species tree\n(IQTree gene trees)") +
  bartheme() +
  theme(legend.position="bottom")
#print(p1)

p2 = ggplot(subset(data, logBF2 > -2 & logBF2 < 50), aes(x=logBF2, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  xlab("log BF2") +
  ylab("") +
  bartheme() +
  theme(legend.position="bottom")
#print(p2)

p3 = plot_grid(p1 + theme(legend.position="none"), p2 + theme(legend.position="none"), ncol=2)
pleg = get_legend(p1)
p = plot_grid(p3, pleg, ncol=1, rel_heights=c(1, .1))
print(p)

```

## BF1 / BF2

```{r bf-ratio-rf, out.width="60%", fig.align = "center", warning=FALSE, fig.height=4}

data$bf.ratio = data$logBF1 / data$logBF2

p1 = ggplot(subset(data, bf.ratio < 10 & bf.ratio > -10), aes(x=bf.ratio, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  #ggtitle("IQtree gene trees") +
  xlab("log BF1 / log BF2") +
  ylab("Weighted RF to species tree\n(IQtree gene trees)") +
  bartheme() +
  theme(legend.position="bottom")
#print(p1)

p2 = ggplot(subset(data, bf.ratio < 5 & bf.ratio > -2), aes(x=bf.ratio, y=iq.wrf, color=acc)) +
  geom_point(size=2, alpha=0.3) +
  #geom_quasirandom(size=2, width=0.25, alpha=0.25, color="#666666") +
  #geom_boxplot(outlier.shape=NA, alpha=0.75, width=0.5, color="#666666") +
  scale_color_manual(values=corecol(numcol=2, pal="wilke")) +
  #labs(color="Accelerated?") +
  #scale_y_continuous(expand=c(0,0)) +
  #ggtitle("") +
  xlab("log BF1 / log BF2") +
  ylab("") +
  bartheme() +
  theme(legend.position="bottom")
#print(p2)

p3 = plot_grid(p1 + theme(legend.position="none"), p2 + theme(legend.position="none"), ncol=2)
pleg = get_legend(p1)
p = plot_grid(p3, pleg, ncol=1, rel_heights=c(1, .1))
print(p)

```


[< Back ](index.html)

















